Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
pf <- read.csv('pseudo_facebook.tsv', sep='\t')
qplot(age,friend_count, data=pf)
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) + 
  geom_point(alpha=1/20, position=position_jitter(height=0), color='orange') + 
  scale_x_continuous(breaks=seq(13,90,2)) + 
  coord_cartesian(xlim=c(13,70), ylim=c(0,1000)) +
  geom_line(stat='summary',fun.y=mean, color='blue') +
  geom_line(stat='summary',fun.y=quantile, prob=.1, color='red', linetype='dashed') +
  geom_line(stat='summary',fun.y=quantile, prob=.5, color='red', linetype='dashed') +
  geom_line(stat='summary',fun.y=quantile, prob=.9, color='red', linetype='dashed')
  
?coord_trans
names(pf)
```



***

### Overplotting
Notes:

```{r Overplotting}

```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}

```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x=age, y=friendships_initiated), data=pf) + 
  geom_point(alpha=1/2, position=position_jitter(height=0)) + 
  coord_trans(y='sqrt') + 
  scale_x_continuous(breaks=seq(13,90,4), limits=c(13,90))
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)
names(pf)
pf.means <- pf %>% 
            group_by(age) %>%
            summarise(mean_friend_count = mean(friend_count))

```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x=age, y=mean_friend_count), data=pf.means) + 
  geom_line()

```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(x=pf[,c('age')], y=pf[,c('friend_count')])
cor.test(x=pf$age, y=pf$friend_count)
with(pf, cor.test(age, friend_count, method='pearson'))
names(pf)
?cor.test
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(                 , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(data=pf, aes(x=www_likes_received, y=likes_received)) + 
  geom_point() + 
  xlim(0, quantile(pf$www_likes_received, 0.95)) + 
  ylim(0, quantile(pf$likes_received, 0.95)) + 
  geom_smooth(method='lm')

subpf <- subset(pf, pf$www_likes_received < quantile(pf$www_likes_received, 0.95) & pf$likes_received < quantile(pf$likes_received, 0.95))
with(pf, cor.test(method='pearson', x=www_likes_received, y=likes_received))



?geom_smooth
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}

```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
data(Mitchell)
?Mitchell
names(Mitchell)

```

Create your plot!

```{r Temp vs Month}
ggplot(data=Mitchell, aes(x=floor(Month/12), y=Temp)) + 
  geom_point() + 
  scale_x_continuous(breaks=seq(0,50,1)) + 
  geom_line(stat='summary', fun.y=quantile, prob=0.5, color='blue') +
  geom_line(stat='summary', fun.y=quantile, prob=0.1, color='red') +
  geom_line(stat='summary', fun.y=quantile, prob=0.9, color='green')


ggplot(data=Mitchell, aes(x=Month %% 12, y=Temp)) + 
  geom_point() + 
  scale_x_continuous(breaks=seq(0,50,1))


ggplot(data=Mitchell, aes(x=Month, y=Temp)) + 
  geom_point() + 
  scale_x_continuous(breaks=seq(0,203,12))

with(Mitchell, cor.test(x=Month, y=Temp))
max(Mitchell$Month)
range(Mitchell$Month)
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
names(pf)
#subset(pf, pf$age == 14 & pf$dob_year != 1999, select=c(dob_month, dob_year))

pf$age_with_months <- pf$age + (12-pf$dob_month)/12

head(pf)

```

***

### Age with Months Means

```{r Age with Months Means}
pf.fc_by_age_months <- pf %>%
                        group_by(age_with_months) %>%
                        summarise(mean_friend_count = mean(friend_count),
                                  median_friend_count = median(friend_count),
                                  n = n()) %>%
                        arrange(age_with_months)

head(pf.fc_by_age_months)

ggplot(data = pf.fc_by_age_months, aes(x=age_with_months, y=mean_friend_count)) + 
  geom_line() + 
  scale_x_continuous(breaks=seq(10,150,1/4)) + 
  coord_cartesian(xlim=c(15,20), ylim=c(250,450))

#mean friend count
#median friend count
#number of users in each group of age_with_months

ggplot(data = pf.fc_by_age_months, aes(x=age_with_months, y=mean_friend_count)) + 
  geom_line() + 
  scale_x_continuous(breaks=seq(13,71,1)) + 
  coord_cartesian(xlim=c(13,71), ylim=c(0,500))


```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

