Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
setwd('/Users/lucasosouza/Desktop/AI/Udacity/R-EDA/lesson5')
pf <- read.delim('../lesson3/pseudo_facebook.tsv')
library(dplyr)
library(tidyr)
library(ggplot2)

summary(pf)
names(pf)

#create a new dataframe called pf.fc_by_age_gender
#contains info on each age AND gender group
#which info? mean_friend_count, median_friend_count, n for number of users

#using dplyr and summarize
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>% 
  summarize(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>%
  ungroup() %>%
  arrange(age)

#ggplot(aes(x = gender, y = age)) + geom_histogram()

```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

library(ggplot2)
ggplot(data=pf.fc_by_age_gender, aes(x=age, y=median_friend_count)) + 
  geom_line(aes(color=gender))

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')

ggplot(pf.fc_by_age_gender.wide, aes(y=(female/male),x=age)) + 
    geom_point() + 
    geom_hline(linetype=2, yintercept=1)

pf$year_joined <- floor(2014 - (pf$tenure/365))
tail(pf$year_joined)
?round
names(pf)

```


***

### Ratio Plot
Notes:

```{r Ratio Plot}

```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}

```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
# PARENTHESIS exclude BRACKETS include
?cut
table(pf$year_joined)
table( cut(pf$year_joined, breaks=c(2004, 2009, 2011, 2012, 2014) ))
pf$year_joined.bucket<- cut(pf$year_joined, breaks=c(2004, 2009, 2011, 2012, 2014))
table(pf$year_joined.bucket)
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}

pf.fc_by_age_gender <- pf %>%
  filter(!is.na(year_joined.bucket)) %>%
  group_by(age, year_joined.bucket) %>% 
  summarize(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>%
  ungroup() %>%
  arrange(age)

ggplot(data=pf.fc_by_age_gender, aes(x=age, y=median_friend_count)) + 
  geom_line(aes(color=year_joined.bucket))

ggplot(data=subset(pf, !is.na(year_joined.bucket)), 
       aes(x=age, y=friend_count))
  + geom_line(stat='summary', 
              fun.y=median, 
              aes(color=year_joined.bucket))

names(pf)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}

# (1) Add another geom_line to code below
# to plot the grand mean of the friend count vs age.
#done

# (2) Exclude any users whose year_joined.bucket is NA.
#done

# (3) Use a different line type for the grand mean.

# As a reminder, the parameter linetype can take the values 0-6:

# 0 = blank, 1 = solid, 2 = dashed
# 3 = dotted, 4 = dotdash, 5 = longdash
# 6 = twodash

ggplot(data=pf.fc_by_age_gender, aes(x=age, y=mean_friend_count)) + 
  geom_line(aes(color=year_joined.bucket)) + 
  geom_line(stat='summary', fun.y=mean, linetype=2)


names(pf)
```

***

### Friending Rate
Notes:

```{r Friending Rate}
#rate - friends by days of tenure - at least 1 day of tenure
with(subset(pf, tenure>0), summary(friend_count/tenure)) #with - very useful

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
# Create a line graph of mean of friendships_initiated per day (of tenure)
# vs. tenure colored by year_joined.bucket.

# You need to make use of the variables tenure,
# friendships_initiated, and year_joined.bucket.

# You also need to subset the data to only consider user with at least
# one day of tenure.

# This assignment is not graded and
# will be marked as correct when you submit.

pf2 <- subset(pf, tenure>0)
pf2$friendships_initiated_per_day <- pf2$friendships_initiated / pf2$tenure
pf2$year_joined <- 2014 - floor(pf2$tenure / 365)
pf2$year_joined.bucket <- cut(pf2$year_joined, breaks=c(2004,2009,2011,2012, 2014))

ggplot(data=pf2, aes(x=tenure,y=friendships_initiated_per_day)) + 
  geom_line(aes(color=year_joined.bucket), stat='summary', fun.y=mean)

ggplot(data=pf2, aes(x=tenure,y=friendships_initiated_per_day)) + 
  geom_line(aes(color=year_joined.bucket))

#smooth chooses the appropriate statistical method based on the data
ggplot(data=pf2, aes(x=tenure,y=friendships_initiated_per_day)) + 
  geom_smooth(aes(color=year_joined.bucket))

#spline
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv('yogurt.csv')
yo$id <- factor(yo$id)
str(yo)

#why change to a factor
summary(yo)

?geom_histogram
?scale_x_continuous

#each line is a purchase
ggplot(yo, aes(x=price)) + 
  geom_histogram(bins=5) + 
  scale_x_continuous(breaks=c(0,100,2))

```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
yo$all_purchases <- yo$strawberry + yo$blueberry + yo$pina.colada + yo$plain + yo$mixed.berry
#another option is to use transform
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
yo <- yo %>% select(-c(all_purchases))

```

***

### Prices over Time
Notes:

```{r Prices over Time}

ggplot(yo, aes(x=time, y=price)) + 
  geom_jitter(alpha=.3, height=2, width=2) #modal - most common prices
?geom_jitter # use geom_jitter to smooth away discretness
?geom_point
?alpha
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(3000)
sample.ids <- sample(levels(yo$id), 16)

p <- ggplot(subset(yo, id %in% sample.ids), aes(x=time, y=price)) + 
  geom_jitter(alpha=.4, height=1, width=1, fill='#0000AA') +
  geom_line() +
  facet_wrap(~id)

ggsave('yogurt-seed3000.png', p)

#At least 4 of the households with greater yogurt consumption show the same pattern - an overall increase in the price over time, but with sharp declines on specific periods. These may or may not refer to sales or usage of coupons.

#The dates of the declines are not the same. If all data is coming from the same physical store, we can imply the price decline is not due to sales, but rather to coupons. Otherwise, if the data is coming from multiple stores, each store could have sales on different days, and we can not infer the cause.

#The remaining households, with lower consumption, shows a steady increase of the price over time.

```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

```{r}
#interesting new library
library(GGally)
themse_set(theme_minimal(20))

set.seed(1836)
pf_subset <- pf[,c[2:15]] #selecting only a few columns
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000)])

```



***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

#oh shit... there is a problem set 5!!
#including another GapMinder analysis
#this is going to take forever